A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2018-11-02 doc/large_strain.tex
fix few more comments

To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent masters thesis produced by a student of the first two of the authors \cite{Mentler2017}.
One possible reason is that the tangent operators\footnote{{\color{red}Here and
below by operators we mean finite-dimensional linear operators on vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
}} (stemming from linearization of the nonlinear balance equations) are
more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether the matrix-free methods can be advantageous in this case.
Another reason could be the use of lower order elements in solid mechanics due to lacking smoothness of the underlying solution.
The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community.
Even higher order elements are rarely employed.
{\color{red}
This can potentially reduce advantage of matrix-free methods, that are generally more competitive for higher order elements.
}
However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.

@@

For the current study, we use the Neo-Hookean model
\begin{gather}
\psi \left( \mathbf{C} \right)
  = \frac{\mu}{2} \left[ \trace{\mathbf{C}} - \trace{\mathbf{I}} - 2 \ln\left( J \right) \right]
  + \lambda \ln^{2}\left( J \right)
\end{gather}
where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively. It can be shown that the first and second derivatives of the strain energy function are given by
\begin{align}
\frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}}
  &= \frac{\mu}{2} \mathbf{I} - \frac{1}{2} \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{C}^{-1} \\
\frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}}
  &= \frac{1}{2}\left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
  + \frac{\lambda}{2} \mathbf{C}^{-1} \otimes \mathbf{C}^{-1}
\end{align}
The Kirchhoff stress and its associated {\color{red}fourth-order material part of spatial tangent tensor} are
\begin{gather}
\boldsymbol{\tau}
  \equiv J \boldsymbol{\sigma}
  = \chi\left( 2 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}} \right)
  = \mu \mathbf{b} - \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{I}
\end{gather}
and
\begin{gather}
J \boldsymbol{\mathcal{C}}
  = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
  + 2 \lambda \mathbf{I} \otimes \mathbf{I}
\end{gather}
where $\boldsymbol{\mathcal{S}}$ is the fourth-order symmetric identity tensor.
The action that $J \boldsymbol{\mathcal{C}}$ performs when contracted with an arbitrary rank-2 symmetric tensor is therefore
\begin{gather}
J \boldsymbol{\mathcal{C}} : \left( \bullet \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)
  + 2 \lambda \trace\left( \bullet \right) \mathbf{I}.
\label{eq:simplified_action}
\end{gather}

>>>>

To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018} and a recent masters thesis produced by a student of the first two of the authors \cite{Mentler2017}.
One possible reason is that the tangent operators\footnote{{\color{red}Here and
below by operators we mean finite-dimensional linear operators on vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
}} (stemming from linearization of the nonlinear balance equations) are
more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
Another reason could be the {\color{red}frequent} use of lower order elements in solid mechanics due to lacking smoothness of the underlying solution.
The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
{\color{red}whereas} higher order elements are rarely employed.
{\color{red}
This can potentially reduce advantage of matrix-free methods, that are generally more competitive for higher order elements.
}
However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.

@@

For the current study, we use the Neo-Hookean model
\begin{gather}
\psi \left( \mathbf{C} \right)
  = \frac{\mu}{2} \left[ \trace{\mathbf{C}} - \trace{\mathbf{I}} - 2 \ln\left( J \right) \right]
  + \lambda \ln^{2}\left( J \right)
\end{gather}
where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively. It can be shown that the first and second derivatives of the strain energy function are given by
\begin{align}
\frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}}
  &= \frac{\mu}{2} \mathbf{I} - \frac{1}{2} \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{C}^{-1} \\
\frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}}
  &= \frac{1}{2}\left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
  + \frac{\lambda}{2} \mathbf{C}^{-1} \otimes \mathbf{C}^{-1}
\end{align}
The Kirchhoff stress and its associated {\color{red}fourth-order material part of spatial tangent tensor} are
\begin{gather}
\boldsymbol{\tau}
  \equiv J \boldsymbol{\sigma}
  = \chi\left( 2 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}} \right)
  = \mu \mathbf{b} - \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{I}
\end{gather}
and
\begin{gather}
J \boldsymbol{\mathcal{C}}
  = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
  + 2 \lambda \mathbf{I} \otimes \mathbf{I}
\end{gather}
where $\boldsymbol{\mathcal{S}}$ is the fourth-order symmetric identity tensor
{\color{red}with $S_{ijkl}=\frac{1}{2}\left[\delta_{ik}\delta_{jl}+\delta_{il}\delta_{jk}\right]$}.
The action that $J \boldsymbol{\mathcal{C}}$ performs when contracted with an arbitrary rank-2 symmetric tensor is therefore
\begin{gather}
J \boldsymbol{\mathcal{C}} : \left( \bullet \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)
  + 2 \lambda \trace\left( \bullet \right) \mathbf{I}.
\label{eq:simplified_action}
\end{gather}
